### 2017-02-03 acknowledgements.tex
Acknowledgements revised

First and foremost, I would like to thank my advisor, Dr. Roman Makarevich for his guidance throughout this project, which would not have been possible without his support, feedback, and constructive criticism.  I would also like to acknowledge the efforts of the other members of my committee, Dr. William Bristow, Dr. Mark Conde, and Dr. Hui Zhang for always providing new ideas and perspectives.  Additionally, I truely appreciate the support and encouragement I received from the entire physics department at UAF, particularly Dr. David Neuman, Dr. Renate Wackerbauer, Dr. Curt Szuberla, Dr. Chung-Sang Ng, Dr. Peter Delamere, Ellen Craig and Jeanie Talbot.  I would like to extend a particularly special thank you to Flora G**** and all the staff of the Keith B. Mather Libray for providing invaluable assistance in retrieving sources for my research, no matter how old and obscure my requests were.

To my family, thank you all for the supporting me for years, particularly my Mom, for making me a much better writer than I have any right to be as a scientist, my Dad, for always being there to answer my programming questions and helping me fix my code when I had now idea how I had broken it, and my brother, Henry, generally for just existing and providing a friendly voice of reason when I really needed it.  Additionally, thank you to my friends Katie, Ari, Bonnie, Heather, Jayne, and Stephanie for being there over these last few years despite being 4000 miles away, to Laura for giving me perspective and a laugh regardless of the time difference.  Finally, I would like to extend a very special thank you to Allison, Rosie, Emma, and Kinley on the homefront for their tireless effort in keeping me sane, healthy, and happy throughout this whole process.  I couldn't have done it without you guys.  Thank you.

>>>>

First and foremost, I would like to thank my advisor, Dr. Roman Makarevich for his guidance throughout this project, which would not have been possible without his continuous support and feedback.  I would also like to acknowledge the efforts of the other members of my committee, Dr. William Bristow, Dr. Mark Conde, and Dr. Hui Zhang for offering new ideas and perspectives.  Additionally, I truely appreciate the support and encouragement I received from the entire physics department at UAF, particularly Dr. David Neuman, Dr. Renate Wackerbauer, Dr. Curt Szuberla, Dr. Chung-Sang Ng, Dr. Peter Delamere, Ellen Craig and Jeanie Talbot, as well as my fellow grad students, Victoriya, Christina, Jason, Robin, Joe, Nathan, Soma, Bishwa, Vitaliy, and John for both the homework and coding assistance and companionship over the last several years.  I would like to extend a particularly special thank you to Flora Grabowska and all the staff of the Keith B. Mather Libray for providing invaluable assistance in retrieving a great number of old, obscure, and difficult to locate sources for my research.

To my family, thank you all for supporting me for years, particularly my Mom, for helping me become a much better writer than I would have been otherwise, my Dad, for always picking up the phone for my programming emergencies, and my brother, Henry, for giving me a friendly voice of reason when I really needed it.  Additionally, I would like to thank my friends back East, Katie, Ari, Bonnie, Heather, Jayne, and Stephanie for supporting me over these last few years despite being 4000 miles away, and to Laura for always brightening my spirits and making my life feel a little less hectic.  Finally, a huge thank you to Allison, Rosie, Emma, and Kinley on the homefront for their tireless effort in keeping me sane, healthy, and happy throughout this whole process.  I couldn't have done it without you guys.  Thank you.

### 2017-02-03 introduction.tex
Acknowledgements revised

\subsection{Polar Patch Formation}
\label{sec:lit_patches}
Large-scale structuring in the polar cap can take a variety of forms, Section \ref{sec:polar_structure}, but here the focus is on large regions of enhanced plasma density known as polar patches  \citep[e.g.][]{Weber1984,Weber1986,Buchau1983,Buchau1985}.  One of the primary interests of this thesis is the relationships between large-scale plasma structures ans small-scale irregularity formation.  This structuring has long been observed around polar patches \citep{Weber1984,Milan2002b,Moen2012}.  Although small-scale structures have also been observed around sun-aligned arcs \citep[e.g.][]{Koustov2012}, these structures also tend to be more complicated and often are surrounded by intense shears and electric fields \citep{Safargaleev2000,Aikio2002,Kozlovsky2007}.  Additionally, polar patches occur relatively frequently, providing a good opportunity to study irregularity coupling at different scales \citep{Rodger1996}.  In theory, many of the results presented in this work pertaining to the strong gradients on the edges of patches could also easily be applied to the edges of polar holes, but to date, far less attention has been paid to polar holes in the literature, making it challenging to accurately put observations in context \citep{Makarevich2015}.

>>>>

\subsection{Polar Patch Formation}
\label{sec:lit_patches}
Large-scale structuring in the polar cap can take a variety of forms, Section \ref{sec:polar_structure}, but here the focus is on large regions of enhanced plasma density known as polar patches  \citep[e.g.][]{Weber1984,Weber1986,Buchau1983,Buchau1985}.  One of the primary interests of this thesis is the relationships between large-scale plasma structures ans small-scale irregularity formation.  This structuring has long been observed around polar patches \citep{Weber1984,Milan2002b,Moen2012}.  Although small-scale structures have also been observed around sun-aligned arcs \citep[e.g.][]{Koustov2012}, these structures also tend to be more complicated and often are surrounded by intense shears and electric fields \citep{Safargaleev2000,Aikio2002,Kozlovsky2007}.  Additionally, polar patches occur relatively frequently, providing a good opportunity to study irregularity coupling at different scales \citep{Rodger1996}.  In theory, many of the results presented in this work pertaining to the strong gradients on the edges of patches could also easily be applied to the edges of polar holes, but to date, far less attention has been paid to polar holes in the literature, making it challenging to accurately put observations in context \citep{Makarevich2015b}.

### 2017-02-03 paper1.tex
Corrections to Abstract and Intro
Formatting corrections made
Acknowledgements revised

This result is significant as it implies a considerable additional source of density as compared with IRI outputs. This density would provide additional refraction and bring backscatter to orthogonality at the ranges consistent with the occurrence model.  This prompts an important question about the origin of this source. An improved agreement between the observed and IRI modeled densities has been recently reported during periods of solar production-dominated dynamics for both URSI and CCIR options \citep{Themens2014}.  The increased errors during periods of little solar-driven production were attributed ot transport processes \citep{Themens2014}.  Following the same idea, a reasonable assumption is that additional density implied by our observations was transported from somewhere else.  Several phenomena may be responsible for this including polar patches \citep{Wever1984}, Sun-aligned arcs \citep{Valladares1994}, and storm-enhanced densities \citep{Foster1993}.  The polar patches in particular are known to travel long distances across the polar cap and even well into the nightside auroral zone \citep{Moen2007,Oksavik2010}.  Some support for the polar patches as a feasible source for the plasma density also comes from the density measurements presented below.

@@

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{p1f6.pdf}
	\caption[Cojugate patchy day example]{Conjugate measurements for two selected events on (a--d) March 14, 2010 and (e-h) April 6, 2010. Shown are the IMF \(B_y\) (blue) and \(B_z\) (red) components (Figures \ref{fig:patchy_example}a and \ref{fig:patchy_example}e), MCM line-of-sight velocity versus UT and slant range in beam 14 (Figures \ref{fig:patchy_example}b and \ref{fig:patchy_example}f), RKN velocity in beam 4 (Figure \ref{fig:patchy_example}c and \ref{fig:patchy_example}g), and RISR-N electron density versus UT and altitude (Figure \ref{fig:patchy_example}d and \ref{fig:patchy_example}h). The color bar is the same for Figures \ref{fig:patchy_example}b, \ref{fig:patchy_example}f, \ref{fig:patchy_example}c, and \ref{fig:patchy_example}g and is given to the right of Figure \ref{fig:patchy_example}f.}
	\label{fig:patchy_example}
\end{figure}

>>>>

This result is significant as it implies a considerable additional source of density as compared with IRI outputs. This density would provide additional refraction and bring backscatter to orthogonality at the ranges consistent with the occurrence model.  This prompts an important question about the origin of this source. An improved agreement between the observed and IRI modeled densities has been recently reported during periods of solar production-dominated dynamics for both URSI and CCIR options \citep{Themens2014}.  The increased errors during periods of little solar-driven production were attributed ot transport processes \citep{Themens2014}.  Following the same idea, a reasonable assumption is that additional density implied by our observations was transported from somewhere else.  Several phenomena may be responsible for this including polar patches \citep{Weber1984}, Sun-aligned arcs \citep{Valladares1994}, and storm-enhanced densities \citep{Foster1993}.  The polar patches in particular are known to travel long distances across the polar cap and even well into the nightside auroral zone \citep{Moen2007,Oksavik2010}.  Some support for the polar patches as a feasible source for the plasma density also comes from the density measurements presented below.

@@

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{p1f6.pdf}
	\caption[An example of conjugate observations of patchy backscatter]{Conjugate measurements for two selected events on (a--d) March 14, 2010 and (e-h) April 6, 2010. Shown are the IMF \(B_y\) (blue) and \(B_z\) (red) components (Figures \ref{fig:patchy_example}a and \ref{fig:patchy_example}e), MCM line-of-sight velocity versus UT and slant range in beam 14 (Figures \ref{fig:patchy_example}b and \ref{fig:patchy_example}f), RKN velocity in beam 4 (Figure \ref{fig:patchy_example}c and \ref{fig:patchy_example}g), and RISR-N electron density versus UT and altitude (Figure \ref{fig:patchy_example}d and \ref{fig:patchy_example}h). The color bar is the same for Figures \ref{fig:patchy_example}b, \ref{fig:patchy_example}f, \ref{fig:patchy_example}c, and \ref{fig:patchy_example}g and is given to the right of Figure \ref{fig:patchy_example}f.}
	\label{fig:patchy_example}
\end{figure}

### 2017-02-03 paper3.tex
Acknowledgements revised

The 2D electric field \(\vec{E}\)  and plasma drift velocity \(\vec{V}_E\) can be found from incoherent scatter radar (ISR) measurements, using the standard ISR method utilizing line-of-sight ion velocity data \citep{Hei08}.  Additionally, ISRs have been previously used to determine density patterns in the ionosphere \citep{Semeter2009,Dahlgren2012a,Dahlgren2012b}. \citet{Dahlgren2012b} qualitatively compared the location of irregularities detected by a coherent HF radar to high-density regions identified by an ISR.  The observations were described in the context of GDI predictions, but the growth rate has not been explicitly evaluated. The current study aims to fill this gap by calculating 2D density gradient fields based on ISR plasma density patterns. Previous studies that evaluated plasma density gradients have found them primarily from in situ measurements from sounding rockets \citep{Moen2012,Lynch2015}, but that method inherently only allows one component of the gradient to be resolved and only allows for measurements along the rocket's flight path.  The new method employed here allows a field of 2D gradient vectors to be calculated, which, together with the 2D electric fields also provided by ISR, allows for experimental investigations of plasma structuring in context of factors \(\uvec{k}\cdot\vec{E}\) and \(\uvec{k}\cdot\uvec{b}\times\vec{G}\) as well their directional equivalents \(\uvec{k}\cdot\uvec{e}\) and \(\uvec{k}\cdot\uvec{b}\times\uvec{g}\), Equation (\ref{eqn:p3gdi}).

@@

This study utilized the northern face of the Incoherent Scatter Radar (ISR) at Resolute Bay, Canada (RISR-N).  This is one of the Advanced Modular Incoherent Scatter Radar (AMISR) systems that consists of an electronically steerable phased array, allowing the radar to make measurements in different directions close to simultaneously by quickly cycling through all beam directions \citep{Bahcivan2010}.  Typically, AMISR systems have two types of pulses, a long pulse for \(F\)-region observations with 72-km range resolution and an alternating code pulse for \(E\)-region observations with 4.5-km range resolution.  The radar measures a power spectrum and the electron density, ion and electron temperatures, and ion drift velocity are found using a standard ISR technique \citep{Evans1969,Risbeth1985,Nicolls2007}.  Calibration accounts for the noise and system constants \citep{Nicolls2007} and is done using either plasma line measurements (for summer daytime) or ionosonde measurements (for all other times) \citep{Bahcivan2010,Themens2014a}.  3D convection velocity and electric field vectors in the $F$ region are also found by combining all ion drift components measured at a particular magnetic latitude (MLAT) and assuming that both \(\vec{E}\) and \(\vec{V}_E\) are constant in magnetic longitude (MLON) \citep{Heinselman2008}.

@@

Figure \ref{fig:radar_map} shows the FoV of both RKN, located at 62.82\(\deg\)N, 93.11\(\deg\)W, geographic (72.6\(\deg\)N, 26.4\(\deg\)W, geomagnetic), and RISR-N, located at 74.73\(\deg\)N, 94.91\(\deg\)W, geographic (82.8\(\deg\)N, 36.8\(\deg\)W geomagnetic).  Geomagnetic coordinates were determined using the Altitude Adjusted Corrected Geomagnetic Coordinates (AACGM) model \citep{Shepard2014}.  The RKN footprint consists of individual radar cells, outlined in dashed lines.  Cells where SNR \(>\) 3 dB are color-coded in SNR measured at 18:16--18:17 UT on November 30, 2011. The locations of the 11 RISR-N beams at 300 km are shown by circles, with the color of each circle representing the electron density, \(n\), measured by RISR-N at 300 km during the post-integration interval closest to that of RKN.

@@

\section{Discussion}
\label{sec:p3discussion}
In this study, small-scale plasma irregularity characteristics were analyzed in the context of factors that affect the linear growth rate of the gradient-drift instability.  Previous experimental studies of GDI and its controlling factors focused on the convection electric field, electron density, fractional perturbation density, and density gradients derived from 1D measurements \citep[e.g.][]{Villian1986,Fukumoto1999,Fukumoto2000,Danskin2002,Oksavik2010,Moen2012,Makarevich2014b,Burston2016}.  In contrast, the current study focused on the previously uninvestigated directional dependence of radar echo occurrence and intensity and comparison with theoretical predictions for arbitrary gradient, electric field, and wave propagation vectors. To accomplish this, 2D density gradient vectors were estimated from multi-point and direct measurements of plasma density by an incoherent scatter radar and, in combination with electric field measurements by an ISR, used to evaluate the GDI growth rate.  In addition, differences between echo occurrence and backscatter power were examined in context of the GDI theory predictions. Below, we discuss three groups of issues aligned with the objectives of this study.

@@

Another factor that can affect both the irregularity occurrence and strength is the background plasma density \(n\) itself. To achieve perpendicularity to the magnetic field in the polar cap, HF radars rely on beam refraction, which is dependent on the plasma density \citep{Bristow2011,Koustov2014,Lamarche2015}.  High plasma density can also reduce the occurrence of backscatter observed by radar by smoothing the gradients required for GDI \citep{Ruohoniemi1997,Koustov2004}, through \(E\)-region shorting \citep{Dan02,Kan12,Lam15}, or through attenuation of the signal through a dense \(D\) region \citep{Danskin2002}.  The peak in occurrence of all echoes in Figure \ref{fig:density}a agrees with previous results that there is an optimal density for HF radar observations of \(F\)-region irregularities \citep{Danskin2002,Makarevich2014b,Koustov2014,Lamarche2015}.

@@

A summary of the results that agree with these predictions is presented in Table \ref{tab:results}.  The relationship between \(\vec{G}\) and \(\vec{V}_E\) described in entries 9 and 10 in Table \ref{tab:results} matches previously-reported observations of an asymmetry between the trailing and leading edge \citep[][]{Weber1984,Milan2002b,Koustov2012,Moe2012}. Most of these previous studies observed asymmetry only qualitatively and compared only the leading and trailing edge of patches instead of considering all directions around it.  The new approach employed here allowed a quantitative assessment using direct estimates from measurements. Apart from results 9 and 10, all other features in Table  \ref{tab:results} have not been reported before.

@@

With respect to the estimates of the growth rates, the highest rates were around 0.002 s\(^{-1}\), Figure \ref{fig:gamma}c, which corresponds to a rise time of \(\sim8\) min.  However, the peak in echo occurrence occurred at \(\gamma\approx0.0001\) s\(^{-1}\), corresponding to a rise time of over 2.5 hours. In comparison with previous studies, \citet{Moe12} found through sounding rocket measurements \(\gamma = 0.021\) s\(^{-1}\) and a rise time of 47.6 s.  \citet{Carlson2007} estimated the rise time when GDI is the only instability mechanism operational as 5--10 min, but they also proposed a 2-step process in which KHI seeds GDI, in which case the rise time would be \(\sim50\) s.  \citet{Burston2016} found that on the edge of polar cap patches, the GDI growth rate is greater than 0.0017 s\(^{-1}\) (10-min rise time) approximately 81\% of the time and greater than 0.016 s\(^{-1}\) (1-min rise time) approximately 11\% of the time. The fastest growth rates estimated in the current study thus agree well with GDI growth rates estimated from previous studies as well as observations of how quickly structuring develops around polar cap patches \citep{Carlson2007}.  Overall, it appears that structuring can be directionally dependent and echo power and occurrence are not necessarily related in a simple way.

>>>>

The 2D electric field \(\vec{E}\)  and plasma drift velocity \(\vec{V}_E\) can be found from incoherent scatter radar (ISR) measurements, using the standard ISR method utilizing line-of-sight ion velocity data \citep{Heinselman2008}.  Additionally, ISRs have been previously used to determine density patterns in the ionosphere \citep{Semeter2009,Dahlgren2012a,Dahlgren2012b}. \citet{Dahlgren2012b} qualitatively compared the location of irregularities detected by a coherent HF radar to high-density regions identified by an ISR.  The observations were described in the context of GDI predictions, but the growth rate has not been explicitly evaluated. The current study aims to fill this gap by calculating 2D density gradient fields based on ISR plasma density patterns. Previous studies that evaluated plasma density gradients have found them primarily from in situ measurements from sounding rockets \citep{Moen2012,Lynch2015}, but that method inherently only allows one component of the gradient to be resolved and only allows for measurements along the rocket's flight path.  The new method employed here allows a field of 2D gradient vectors to be calculated, which, together with the 2D electric fields also provided by ISR, allows for experimental investigations of plasma structuring in context of factors \(\uvec{k}\cdot\vec{E}\) and \(\uvec{k}\cdot\uvec{b}\times\vec{G}\) as well their directional equivalents \(\uvec{k}\cdot\uvec{e}\) and \(\uvec{k}\cdot\uvec{b}\times\uvec{g}\), Equation (\ref{eqn:p3gdi}).

@@

This study utilized the northern face of the Incoherent Scatter Radar (ISR) at Resolute Bay, Canada (RISR-N).  This is one of the Advanced Modular Incoherent Scatter Radar (AMISR) systems that consists of an electronically steerable phased array, allowing the radar to make measurements in different directions close to simultaneously by quickly cycling through all beam directions \citep{Bahcivan2010}.  Typically, AMISR systems have two types of pulses, a long pulse for \(F\)-region observations with 72-km range resolution and an alternating code pulse for \(E\)-region observations with 4.5-km range resolution.  The radar measures a power spectrum and the electron density, ion and electron temperatures, and ion drift velocity are found using a standard ISR technique \citep{Evans1969,Rishbeth1985,Nicolls2007a}.  Calibration accounts for the noise and system constants \citep{Nicolls2007a} and is done using either plasma line measurements (for summer daytime) or ionosonde measurements (for all other times) \citep{Bahcivan2010,Themens2014}.  3D convection velocity and electric field vectors in the $F$ region are also found by combining all ion drift components measured at a particular magnetic latitude (MLAT) and assuming that both \(\vec{E}\) and \(\vec{V}_E\) are constant in magnetic longitude (MLON) \citep{Heinselman2008}.

@@

Figure \ref{fig:radar_map} shows the FoV of both RKN, located at 62.82\(\deg\)N, 93.11\(\deg\)W, geographic (72.6\(\deg\)N, 26.4\(\deg\)W, geomagnetic), and RISR-N, located at 74.73\(\deg\)N, 94.91\(\deg\)W, geographic (82.8\(\deg\)N, 36.8\(\deg\)W geomagnetic).  Geomagnetic coordinates were determined using the Altitude Adjusted Corrected Geomagnetic Coordinates (AACGM) model \citep{Shepherd2014}.  The RKN footprint consists of individual radar cells, outlined in dashed lines.  Cells where SNR \(>\) 3 dB are color-coded in SNR measured at 18:16--18:17 UT on November 30, 2011. The locations of the 11 RISR-N beams at 300 km are shown by circles, with the color of each circle representing the electron density, \(n\), measured by RISR-N at 300 km during the post-integration interval closest to that of RKN.

@@

\section{Discussion}
\label{sec:p3discussion}
In this study, small-scale plasma irregularity characteristics were analyzed in the context of factors that affect the linear growth rate of the gradient-drift instability.  Previous experimental studies of GDI and its controlling factors focused on the convection electric field, electron density, fractional perturbation density, and density gradients derived from 1D measurements \citep[e.g.][]{Villain1986,Fukumoto1999,Fukumoto2000,Danskin2002,Oksavik2010,Moen2012,Makarevich2014b,Burston2016}.  In contrast, the current study focused on the previously uninvestigated directional dependence of radar echo occurrence and intensity and comparison with theoretical predictions for arbitrary gradient, electric field, and wave propagation vectors. To accomplish this, 2D density gradient vectors were estimated from multi-point and direct measurements of plasma density by an incoherent scatter radar and, in combination with electric field measurements by an ISR, used to evaluate the GDI growth rate.  In addition, differences between echo occurrence and backscatter power were examined in context of the GDI theory predictions. Below, we discuss three groups of issues aligned with the objectives of this study.

@@

Another factor that can affect both the irregularity occurrence and strength is the background plasma density \(n\) itself. To achieve perpendicularity to the magnetic field in the polar cap, HF radars rely on beam refraction, which is dependent on the plasma density \citep{Bristow2011,Koustov2014,Lamarche2015}.  High plasma density can also reduce the occurrence of backscatter observed by radar by smoothing the gradients required for GDI \citep{Ruohoniemi1997,Koustov2004}, through \(E\)-region shorting \citep{Danskin2002,Kane2012,Lamarche2015}, or through attenuation of the signal through a dense \(D\) region \citep{Danskin2002}.  The peak in occurrence of all echoes in Figure \ref{fig:density}a agrees with previous results that there is an optimal density for HF radar observations of \(F\)-region irregularities \citep{Danskin2002,Makarevich2014b,Koustov2014,Lamarche2015}.

@@

A summary of the results that agree with these predictions is presented in Table \ref{tab:results}.  The relationship between \(\vec{G}\) and \(\vec{V}_E\) described in entries 9 and 10 in Table \ref{tab:results} matches previously-reported observations of an asymmetry between the trailing and leading edge \citep[][]{Weber1984,Milan2002b,Koustov2012,Moen2012}. Most of these previous studies observed asymmetry only qualitatively and compared only the leading and trailing edge of patches instead of considering all directions around it.  The new approach employed here allowed a quantitative assessment using direct estimates from measurements. Apart from results 9 and 10, all other features in Table  \ref{tab:results} have not been reported before.

@@

With respect to the estimates of the growth rates, the highest rates were around 0.002 s\(^{-1}\), Figure \ref{fig:gamma}c, which corresponds to a rise time of \(\sim8\) min.  However, the peak in echo occurrence occurred at \(\gamma\approx0.0001\) s\(^{-1}\), corresponding to a rise time of over 2.5 hours. In comparison with previous studies, \citet{Moen2012} found through sounding rocket measurements \(\gamma = 0.021\) s\(^{-1}\) and a rise time of 47.6 s.  \citet{Carlson2007} estimated the rise time when GDI is the only instability mechanism operational as 5--10 min, but they also proposed a 2-step process in which KHI seeds GDI, in which case the rise time would be \(\sim50\) s.  \citet{Burston2016} found that on the edge of polar cap patches, the GDI growth rate is greater than 0.0017 s\(^{-1}\) (10-min rise time) approximately 81\% of the time and greater than 0.016 s\(^{-1}\) (1-min rise time) approximately 11\% of the time. The fastest growth rates estimated in the current study thus agree well with GDI growth rates estimated from previous studies as well as observations of how quickly structuring develops around polar cap patches \citep{Carlson2007}.  Overall, it appears that structuring can be directionally dependent and echo power and occurrence are not necessarily related in a simple way.
